---
title: "Bayesian Cognitive Modeling"
subtitle: "Tutorial"
author: "Linus Hof, Nuno Busch"
date: "`r Sys.time()`"
format: 
  html:
    keep-md: true
editor_options: 
  markdown: 
    wrap: sentence
  chunk_output_type: console
---

```{r include=FALSE}
# packages 
pacman::p_load(readxl , 
               tidyverse , 
               magrittr , 
               knitr, 
               here
               )

set.seed(1) # setting seed to make random samples reproducible
```

```{r include=FALSE}
here::i_am("tutorials/BayesianCognitiveModeling/bayesian_cognitive_modeling.Rmd")
```


## Risky Choice

Sometimes in your life, you make choices between certain outcomes (e.g., what to eat from your fridge). In other situations, you have to make risky choices:
When making decisions under risk, choosing an option does not always lead to the same outcome.
Instead, choosing an option can lead to one of several outcomes with some probability. A classical example would be playing a lottery, another real life example would be deciding to buy insurance or not (where the future event of getting a disease is not for sure but may only happen with some predicted probability).

In our studies, we often investigate people's choice behavior between lotteries ("risky prospects").
Consider the following risky choice problem from a study of [Kellen et al. (2016)](https://linkinghub.elsevier.com/retrieve/pii/S0010027716302104), involving two options, $A$ and $B$. 
The different columns represent different features of both options.

```{r}
#| echo: false
#| output: false

# read data
problems <- read_xlsx(here("tutorials", "BayesianCognitiveModeling", "kellen2016.xlsx"), sheet = 1) # problem features
description <- read_xlsx(here("tutorials", "BayesianCognitiveModeling", "kellen2016.xlsx"), sheet = 3, col_names = TRUE) # DFD choices 

# clean data

problems %<>% 
  select(-c(`...7`,`...8`,`...13`,`...14`)) %>% 
  mutate(domain = case_when(outcomeA1 >= 0 & outcomeA2 >= 0 & outcomeB1 >= 0 & outcomeB2 >= 0 ~ "Gain" , 
                            outcomeA1 <= 0 & outcomeA2 <= 0 & outcomeB1 <= 0 & outcomeB2 <= 0 ~ "Loss" , 
                            .default = "Mixed"))  %>% 
  select( problem, label, domain, everything()) %>% 
  arrange(domain)

description  %<>% rename(label = ...1)
```



```{r}
#| eval: false

# read data
problems <- read_xlsx("tutorials/BayesianCognitiveModeling/kellen2016.xlsx", sheet=1) # problem features

# clean data
problems %<>% 
  select(-c(`...7`,`...8`,`...13`,`...14`)) %>% 
  mutate(domain = case_when(outcomeA1 >= 0 & outcomeA2 >= 0 & outcomeB1 >= 0 & outcomeB2 >= 0 ~ "Gain" , 
                            outcomeA1 <= 0 & outcomeA2 <= 0 & outcomeB1 <= 0 & outcomeB2 <= 0 ~ "Loss" , 
                            .default = "Mixed"))  %>% 
  select( problem, label, domain, everything()) %>% 
  arrange(domain)
```


```{r}
kable(problems[1,])
```

For instance, `Option B` has two possible outcomes, indicated by `outcomeB1` and `outcomeB2`. 
When choosing this option, one of the outcomes is obtained.
Specifically, `outcomeB1`=`r problems[[1,'outcomeB1']]` occurs with `probB1``r problems[[1,'probB1']]` and `outcomeB2`=`r problems[[1,'outcomeB2']]` occurs with `probB2=``r problems[[1,'probB2']]`. 
When choosing this option repeatedly, one therefore expects to obtain `4` most of the time, but sometimes also a `0`.
To illustrate, we simulate choosing `Option B` 100 times. 

```{r}
sample(c(problems[[1,'outcomeB1']], problems[[1,'outcomeB2']]) , 
       size=100 ,
       replace = TRUE ,
       prob=c(problems[[1,'probB1']], problems[[1,'probB2']]))
```

Turning to `Option A`, we notice that `outcomeA1`=`r problems[[1,'outcomeA1']]` occurs with `probA1`=`r problems[[1,'probA1']]` and `outcomeA2`=`r problems[[1,'outcomeA2']]` occurs with `probA2`=`r problems[[1,'probA2']]`.
Thus, this option is effectively a safe option, where `outcomeA2` does not really exist and we instead always obtain a `3`.
To illustrate, we also simulate choosing `Option A` 100 times. 

(Note) that we still need to keep the (in reality non-existent) outcome A2 in our dataframe for modeling purposes. We will get to that later.)

```{r}
sample(c(problems[[1,'outcomeA1']], problems[[1,'outcomeA2']]) , 
       size=100 ,
       replace = TRUE ,
       prob=c(problems[[1,'probA1']], problems[[1,'probA2']]))
```

In a typical risky choice study, people are presented with the possible outcomes and probabilities of both options and need to decide which of the options they prefer (decisions from description). 
However, most studies involve more than just one choice problem. 
For instance, in the study by Kellen et al. (2016), participants were presented with `r nrow(problems)` different choice problems, each represented by a row in the table below. 
Each problem is a unique combinations of the outcomes and probabilities for `Option A` and `Option B`.

The column `domain` indicates whether all possible outcomes are $\geq 0$ (`Gain`), or whether all outcome are $\leq 0$ (`Loss`), or whether the problem involves some outcomes that are $>0$ and some that are a $<0$ - i.e., a problem featuring a gamble that could lead to either a gain, or a loss. 

```{r}
kable(problems)
```


## Expected Value Maximization

What do you think, how do people choose between such lotteries? They are probably trying to maximize rewards (gain the highest outcome values and avoid losing any) on the long run, right?

This assumption is reflected in an influential theory: According to this view, the rational way to make such decisions under risk is to choose the option with the highest expected value (EV), i.e., 

$$
\arg\max  EV = \sum_i^np_ix_i \; ,
$$
where $x_i$ are the possible outcome of the option and $p_i$ are their respective probabilities.
That is, for each option, people should first multiple all outcomes by their probabilities and then sum up the products to obtain the EV.
Then they should choose the option with the higher EV.
The following code follows this procedure for all choice problems used by Kellen et al. (2016).
Below we showcase the choice predicted by EV maximization for the first five gambles of the dataframe, where `max_ev_det` refers to the lottery with the higher EV (gamble A = 0, gamble B = 1) - i.e., the predicted choice.


```{r}
problems %<>%  
  mutate(ev_A = round( probA1*outcomeA1 + probA2*outcomeA2 , 2) , # computes EV of option A and rounds the result to 2 decimal places
         ev_B = round( probB1*outcomeB1 + probB2*outcomeB2 , 2) , # computes EV of option B and rounds the result to 2 decimal places
         # choose option A (coded as 0) when ev_A > ev_B, otherwise B (coded as 1): nothing when ev_A = ev_B
         max_ev_det = case_when(ev_A > ev_B ~ 0 , # choose option A when ev_A > b
                                ev_A < ev_B ~ 1 , # choose option B when ev_A < b
                                ev_A == ev_B ~ NA ) # choose nothing when EV's are equal (NA refers to missing values that are created here)
         )
kable(head(problems))
```

One of the shortcomings of the idea of EV maximization is that people do not seem maximize expected values in real life. 
Kellen et al. (2016) collected the choices of `r ncol(description)-1` participants on all of the `r nrow(description)` choice problems from above.
To illustrate, the following table shows the choices of 10 participants (columns) on the first 10 choice problems (rows). 


```{r}
#| echo: false
#| output: false

# read and clean data
description <- read_xlsx(here("tutorials", "BayesianCognitiveModeling", "kellen2016.xlsx"), sheet = 3, col_names = TRUE)
description  %<>% rename(label = ...1)
```

```{r}
#| eval: false

# read and clean data
description <- read_xlsx("tutorials/BayesianCognitiveModeling/kellen2016.xlsx", sheet = 3, col_names = TRUE)
description  %<>% rename(label = ...1)
```

```{r}
kable(description[1:10, 1:11])
```

Since we know for all problems whether `Option A` (`1`) or `Option B` (`0`) has the higher EV, we can compute the proportion of choices, in which participants chose the option with the higher EV. 

The last column of each row, `ev_max_ob`, shows the proportion of people which choose the option with the higher EV in the respective problem.
It is easy to see that people often deviate from strict EV maximization, sometimes very systematically.


```{r}
problems <- problems %>%   
  left_join(description, by = join_by(label)) %>% # here, we merge our choice data with the dataframe containing the problem descriptions and the calculated EV and choice predictions.
  mutate(ev_max_obs = round ( case_when(max_ev_det == 0 ~ rowMeans(select(., `1_er171989`:`99_AA211989`), na.rm = TRUE)  , # this command computes the proportion of participants that maximized EV in a particular choice problem.
                                        max_ev_det == 1 ~ 1 - rowMeans(select(., `1_er171989`:`99_AA211989`), na.rm = TRUE) 
                                        ) , 
                              )
         ) 

problems %>% 
  select(problem:max_ev_det, ev_max_obs) %>%
  kable()
```



## Cumulative Prospect Theory: A Cognitive Model for Risky Choice

One of the key accomplishments in decision making research is the development of cumulative prospect theory (CPT) by Amos Tversky and Daniel Kahneman, for which Kahneman received the Nobel prize in 2002. Prospect theory aims to describe what choices people are actually making, if not maximizing the EV. 

CPT retains the fundamental idea of EV maximization that people weight the outcomes of each option by their probability. 
However, a core innovation is the assumption that people do not treat the outcomes and probabilities objectively as they are, but instead represent them subjectively in a way that they become distorted. 

We will explain the formalities in detail below, but the main idea may be illustrated with simplifying examples:  
**Subjective utilities:** Outcomes may not have the same subjective value for everyone, dependent on a person's current financial status (the reference point), and desirability may also change depending on the outcome. For example, 10 EUR may feel subjectively more valuable to you when you get them as a bonus on top of 5 EUR (total of 15 EUR), than when you get them as bonus on top of 100 EUR (= total of 110). This is called diminishing marginal utility.  
**Loss aversion**: The concept of loss aversion complements that of subjective utilities by the observation that "losses loom larger than gains": For example, most people would prefer sticking with the status quo instead of playing when offered a toin coss with equal probility of winning 120 EUR or losing 100 EUR, despite positive expected value. Consequently losses are given more weight than equally sized gains.
**Subjective decision weights:** Individuals may behave as if some events are more or less likely than they actually are. For example, people typically attribute excessive weight to events with low probability and insufficient weight to events with high probability (Do you think this could explain why people go to casinos?).

In other words, prospect theory assumes that people choose the option that yields the higher expected utility for them, where the expected utility of an option is determined by summing up the subjective utility of each outcome weighted by its decision weight.

Mathematically, these assumptions are expressed by replacing the outcome values $x_i$ with subjective values (subjective utilities) $v(x_i)$ and replacing the probabilities with subjective decision weight $\pi_i$.

$$
\arg\max_x V = \sum_i^n v(x_i) \pi_i \; ,
$$

where $v(.)$ is a so-called value function that creates subjective utilities by distorting objective outcome values. $\pi_i$ on the other hand is obtained by transforming the objective probabilities into decision weights with a probability weighting function $w(.)$.

Next, we consider the value and the probability weighting function in more detail.

### Value function

The value function takes each objective outcome $x_i$ as input and returns a respective subjective value $v(x_i)$ as output, according to:

$$
v(x_i) = \begin{cases} 
x_i^\alpha & x_i \geq 0 \; ,\\
-\lambda |x_i|^\alpha & x_i <0 \
\end{cases}
$$
where $\alpha$ is a number that usually takes values between $0$ and $1$, and $\lambda$ is a number typically $>0$. Both $\alpha$ and $\lambda$ are called *free parameters* because they can take varying values which can be estimated: For example, given choice data from a certain participant, we can estimate what parameters fit best to their value function to best explain the why they chose what they chose (we'll get to that later).  

$\alpha$ controls the curvature of the value function (the diminishing marginal utility), $\lambda$ controls the steepness of the value function in the loss domain relative to the gain domain (loss aversion).
The figure below illustrates the form of this value transformation when $\lambda = 1.2$, for different values of alpha: $\alpha = 1$, $\alpha = .9$, $\alpha = .8$.

```{r}
vf <- expand_grid(outcome = c(seq(-10, 10, .1)) , 
                  alpha = seq(from = .8, to = 1, by = .1)) %>% 
  mutate(v = case_when(outcome >= 0 ~ outcome^alpha , 
                       outcome < 0 ~ -1.2*(abs(outcome)^alpha) 
                       )
         ) 

vf %>% ggplot(aes(outcome, v, group = as.factor(alpha), color=as.factor(alpha))) + 
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  geom_abline(slope=1, linetype='dashed', linewidth = 1, color = 'gray') + 
  geom_line(linewidth = 1) + 
  theme_minimal()
```

It can be seen that in the case of $\alpha = 1$ (blue line), all subjective values $v(x)$ are equal the objective outcomes $x$ (except for $x_i < 0$: because of $\lambda > 1$. 
The smaller $\alpha$ gets, the more curved (concave) is the line, indicating that the same change in objective outcomes (on the x-axis) leads to a smaller change in the subjective values (on the y-axis), the more extreme (more positive/negative) the objective outcomes get.
This is the most common form observed in empirical data when fitting the $\alpha$ parameter to the observed choices (see below).  

### Probability weighting function

The weighting function takes a probability $p_i$ between $0$ and $1$ as input and returns a *transformed* probability (a decision weight) $w(p_i)$ as output, according to:

$$
w(p_i) = \frac{\delta p_i^\gamma}{\delta p_i^\gamma + (1-p_i)^\gamma} \; ,
$$
where $\gamma$ is a number that usually takes values between $0$ and $2$ and $\delta$ is a number that usually takes values between $0$ and $5$. 

The figure below illustrates the form of this transformation when $\gamma = .5$, $\gamma = .1$ or $\gamma = 1.5$ (line colors), and when $\delta = .4$, $\delta = 1.2$ or $\delta=2$ (different plot facets).

```{r}
#| fig-width: 10
#| fig-height: 3

wf <- expand_grid(p = c(seq(0, 1, .01)) , 
                  gamma = c(.5, 1, 1.5) , 
                  delta = c(.4, 1, 2)) %>%
  mutate(w = ( delta*(p^gamma) ) / (  (delta*(p^gamma)) + (1-p)^gamma ) )


wf %>% ggplot(aes(p, w, group = as.factor(gamma), color = as.factor(gamma))) +
  facet_wrap(~delta, nrow=1) + 
  geom_abline(slope=1, linetype='dashed', linewidth = 1, color = 'gray') +
  geom_line(linewidth = 1) + 
  theme_minimal()
```

It can be seen that in the case of $\gamma = \delta =  1$ (green line, middle panel), all decision weights $w(p)$ are equal to the objective probabilities $p$. 
However, when $\gamma \neq 1$ and/or $\delta \neq 1$, the function gets curved, indicating a nonlinear transformation of probabilities.  

The exact form of the transformation depends on the combination of $\gamma$ and $\delta$. 
Generally, the function takes an inverse S-shape when  $\gamma < 1$ (red line), but an S-shape when $\gamma > 1$ (red line). 
$\gamma$ is therefore often referred to as the curvature parameter and reflects probability sensitivity - how accurately objective probabilities are represented.
When $\delta = 1$, the function intersects the diagonal at the probability midpoint $p=.5$. The function is more elevated (runs above the diagonal most of the time) when $\delta > 1$, but less elevated (runs below the diagonal most of the time) when $\delta < 1$. Consequently, $\delta$ is the *elevation parameter* and is sometimes psychologically interpreted as optimism or pessimism.  

In CPT, the transformations of the probability weighting function are used to derive the decisions weights $\pi$. However, in some applications, the function might differ depending on whether a lottery contains losses or gains (more on that later).

When each option has at most two different outcomes that are all in the gain domain, the decisions weights are derived according to:  

$$
\begin{align}
\pi_{high}^+ &= w(p_{high}) \\
\pi_{low}^+ &= 1 - \pi_{high}^+ 
\end{align}
$$
When each option has at most two different outcomes that are all in the loss domain, the decisions weights are derived according to:   

$$
\begin{align}
\pi_{low}^- &= w(p_{low}) \\
\pi_{high}^- &= 1 - \pi_{low}^+ 
\end{align}
$$
The subjective values and decision weights that are obtained via these two transformations are then used to compute the subjective valuations of each option. 
Similar to EV maximization, it is assumed that the option with the highest valuation is chosen. 

Here, we first rank-order the outcomes and their probabilities according to the magnitude of the outcomes.

```{r}
cpt <- problems %>% 
  mutate( # rank order attributes
    hA = case_when(outcomeA1 > outcomeA2 ~ outcomeA1, 
                   outcomeA1 < outcomeA2 ~ outcomeA2, 
                   outcomeA1 == outcomeA2 ~ outcomeA1) , 
    lA = case_when(outcomeA1 > outcomeA2 ~ outcomeA2, 
                   outcomeA1 < outcomeA2 ~ outcomeA1,
                   outcomeA1 == outcomeA2 ~ outcomeA2) ,
    hB = case_when(outcomeB1 > outcomeB2 ~ outcomeB1, 
                   outcomeB1 < outcomeB2 ~ outcomeB2, 
                   outcomeB1 == outcomeB2 ~ outcomeB1) , 
    lB = case_when(outcomeB1 > outcomeB2 ~ outcomeB2, 
                   outcomeB1 < outcomeB2 ~ outcomeB1,
                   outcomeB1 == outcomeB2 ~ outcomeB2) ,
    p_hA = case_when(outcomeA1 > outcomeA2 ~ probA1,
                     outcomeA1 < outcomeA2 ~ probA2, 
                     outcomeA1 == outcomeA2 ~ probA1) , 
    p_lA = case_when(outcomeA1 > outcomeA2 ~ probA2,
                     outcomeA1 < outcomeA2 ~ probA1,
                     outcomeA1 == outcomeA2 ~ probA2) ,
    p_hB = case_when(outcomeB1 > outcomeB2 ~ probB1,
                     outcomeB1 < outcomeB2 ~ probB2,
                     outcomeB1 == outcomeB2 ~ probB1) , 
    p_lB = case_when(outcomeB1 > outcomeB2 ~ probB2, 
                     outcomeB1 < outcomeB2 ~ probB1,
                     outcomeB1 == outcomeB2 ~ probB2)) %>%
  select(!(probA1:outcomeB2)) %>% 
  select(problem, label, domain, hA:p_lB, ev_A, ev_B, max_ev_det, ev_max_obs, everything()) 
```



```{r}
#| echo: false

cpt_problems <- cpt %>% select(problem:p_lB)
cpt_choices <- cpt %>% select(`1_er171989`:`99_AA211989`)
```

We then determine what choice CPT predicts for each choice problem given a set of parameter values that we can set. Here, we illustrate a case where objective values and probabilities are not distorted: all parameters equal 1. Based on these parameters, CPT makes the same predictions as EV maximization.

```{r}

lambda <- 1
alpha <- 1
gamma <- 1
delta <- 1

cpt_pred <- cpt %>%
  
  mutate(
    
    ### transform attributes
    
    # value function: transformation of outcome values
    v_hA = if_else(hA >= 0, hA^alpha ,  -lambda*(abs(hA)^alpha)) , 
    v_lA = if_else(lA >= 0, lA^alpha ,  -lambda*(abs(lA)^alpha)) , 
    v_hB = if_else(hB >= 0, hB^alpha ,  -lambda*(abs(hB)^alpha)) ,
    v_lB = if_else(lB >= 0, lB^alpha ,  -lambda*(abs(lB)^alpha)) , 
    
    # probability weighting function: transformation of probabilities
    pi_hA = case_when(hA >= 0 ~ ( delta*(p_hA^gamma) ) / (  (delta*(p_hA^gamma)) + p_lA^gamma ) , 
                      hA < 0 ~ 1 - ( delta*(p_lA^gamma) ) / (  (delta*(p_lA^gamma)) + p_hA^gamma ) ) ,
    pi_lA = 1 - pi_hA , 
    
    pi_hB = case_when(hB >= 0 ~ ( delta*(p_hB^gamma) ) / (  (delta*(p_hB^gamma)) + p_lB^gamma ) , 
                      hB < 0 ~ 1 - ( delta*(p_lB^gamma) ) / (  (delta*(p_lB^gamma)) + p_hB^gamma )
    ) ,  
    pi_lB = 1 - pi_hB ,
    
    ### option valuation and choice
    
    # Calculates the subjective valuation of each option in a choice problem.
    cpt_A = pi_hA*v_hA + pi_lA*v_lA ,   
    cpt_B = pi_hB*v_hB + pi_lB*v_lB ,
    
    # Compare option valuations and deterministically predict a choice based on this comparison.
    cpt_det = case_when(cpt_A > cpt_B ~ 0 , 
                        cpt_A < cpt_B ~ 1 , 
                        cpt_A == cpt_B ~ NA )
    ) 
    
# here, we calculate the proportion of people whose choice aligns with CPT's predictions.
test <- cpt_pred %>% 
  mutate(cpt_max_obs = round ( case_when(cpt_det == 0 ~ rowMeans(select(., `1_er171989`:`99_AA211989`), na.rm = TRUE)  , 
                                         cpt_det == 1 ~ 1-rowMeans(select(., `1_er171989`:`99_AA211989`), na.rm = TRUE) 
                                        ) , 2
                              )
         ) %>%
  select(problem, label, domain, ev_max_obs, cpt_max_obs) %>% 
  mutate(equal = ev_max_obs==cpt_max_obs)

# here, we output the mean of empirical choices aligning with EV maximization and with CPT predictions, respectively.
mean(test$ev_max_obs, na.rm=T) 
mean(test$cpt_max_obs, na.rm=T) 
```

If, however, we choose different parameters that will yield curved value and weighting functions; the CPT predictions differ from EV maximization.
Here, we show this based on an empirically plausible set of parameter values and calculate the proportion of people for each choice problem in the empirical data from Kellen et al. whose choice aligned with CPT's predictions:

```{r}

lambda <- 1.2
alpha <- 0.8
gamma <- 0.6
delta <- 1

cpt_pred <- cpt %>%
  
  mutate(
    
    ### transform attributes
    
    # value function: transformation of outcome values
    v_hA = if_else(hA >= 0, hA^alpha ,  -lambda*(abs(hA)^alpha)) , 
    v_lA = if_else(lA >= 0, lA^alpha ,  -lambda*(abs(lA)^alpha)) , 
    v_hB = if_else(hB >= 0, hB^alpha ,  -lambda*(abs(hB)^alpha)) ,
    v_lB = if_else(lB >= 0, lB^alpha ,  -lambda*(abs(lB)^alpha)) , 
    
    # probability weighting function: transformation of probabilities
    pi_hA = case_when(hA >= 0 ~ ( delta*(p_hA^gamma) ) / (  (delta*(p_hA^gamma)) + p_lA^gamma ) , 
                      hA < 0 ~ 1 - ( delta*(p_lA^gamma) ) / (  (delta*(p_lA^gamma)) + p_hA^gamma ) ) ,
    pi_lA = 1 - pi_hA , 
    
    pi_hB = case_when(hB >= 0 ~ ( delta*(p_hB^gamma) ) / (  (delta*(p_hB^gamma)) + p_lB^gamma ) , 
                      hB < 0 ~ 1 - ( delta*(p_lB^gamma) ) / (  (delta*(p_lB^gamma)) + p_hB^gamma )
    ) ,  
    pi_lB = 1 - pi_hB ,
    
    ### option valuation and choice
    
    # Calculates the subjective valuation of each option in a choice problem.
    cpt_A = pi_hA*v_hA + pi_lA*v_lA ,   
    cpt_B = pi_hB*v_hB + pi_lB*v_lB ,
    
    # Compare option valuations and deterministically predict a choice based on this comparison.
    cpt_det = case_when(cpt_A > cpt_B ~ 0 , 
                        cpt_A < cpt_B ~ 1 , 
                        cpt_A == cpt_B ~ NA )
    ) 
    
# here, we calculate the proportion of people whose choice aligns with CPT's predictions.
test <- cpt_pred %>% 
  mutate(cpt_max_obs = round ( case_when(cpt_det == 0 ~ rowMeans(select(., `1_er171989`:`99_AA211989`), na.rm = TRUE)  , 
                                         cpt_det == 1 ~ 1-rowMeans(select(., `1_er171989`:`99_AA211989`), na.rm = TRUE) 
                                        ) , 2
                              )
         ) %>%
  select(problem, label, domain, ev_max_obs, cpt_max_obs) %>% 
  mutate(equal = ev_max_obs==cpt_max_obs)


# here, we output the mean of empirical choices aligning with EV maximization and with CPT predictions, respectively.
mean(test$ev_max_obs, na.rm=T) 
mean(test$cpt_max_obs, na.rm=T) 
```

Comparing the mean of choices that align with EV maximization vs. CPT, we see that CPT better predicts our empirical choices given the chosen set of parameter values. Could this prediction accuracy be improved even further if we chose other parameter values? How can we find out which parameter values fit best?

## Fitting CPT

In the above demonstration of CPT, we showed that applying the transformations implemented in CPT lead to an improved prediction over EV maximization. 
The behavior of the curves for different parameter values constrains the plausible (combinations of) parameter values to some degree - but in the previous examples, we nonetheless set their values in a rather unsystematic way. 
Hence, it is likely that another combination of parameters might yield a better prediction.  


But how to find the best fitting combination of parameter values $\Theta$ for our model? 
In other words, how do we get from our initial beliefs about the plausibility of the parameter values, $p(\Theta) = p(\alpha, \lambda, \gamma, \delta)$, (before having seen the data), to updated "beliefs" about the parameters values after having seen the data $p(\Theta|D)$? 

$$
p(\Theta) \to p(\Theta|D) \; \text{?}
$$
Here is the intuition behind the solution for this updating problem from a *Bayesian* perspective:

1. Take a set of parameter values for the parameters in your model
2. Calculate the probability of the data given these parameter values (likelihood) 
3. Weigh (multiply) the likelihood with the initial plausibility of parameter values (the prior)

The product we obtain from Step 3 (i.e., the multiplication of prior and likelihood) gives us the updated probability of the respective parameter values given the data. This is called the *posterior probability*: 

$$
p(\Theta|D) \propto p(\Theta) p(D|\Theta)
$$ 
If we repeat these steps for all possible combinations of parameter values, we can simply take the parameter values that have the biggest posterior probability given the data.
Before demonstrating how we can compute these posteriors for our CPT model using `R` and a simulation program calls `JAGS`, we explain the different parts of this Bayesian updating process in more detail.

### Prior

The prior probability, $p(\Theta)$ expresses our initial beliefs about the plausibility of different parameter values. For example, before testing if a coin (or dice) is biased, you would probably have a prior belief that the coin has a .50 probability of landing on each side (1/6 = .17 for each side of the classical dice). Additionally, you will probably find it unlikely that any side has a probability of >.90 of landing on it, even if the coin or dice is biased (that would defy the law of physics if coin landed on heads if you let it fall to the ground with heads up from a height of 10 cm). 
The main function of the prior is to indicate which parameter values could be obtained in principle (the possible range of values), and how plausible the values on this range are relatively to each other.
In other words, the prior is a probability distribution over parameter values. 
<!--maybe we can get more detailed here, e.g., provide an example-->

### Likelihood

#### The Most Simple Choice Model

The likelihood is the probability of the data given the model and a set of parameter values.

To illustrate, a much simpler model than CPT for the risky choice problems from Kellen et al. (2016) would be a binomial model which predicts the choice of `Option 1` over `Option 2` with a constant probability $\theta$.
That is, when $\theta=1$, it always predicts choosing `Option 1`. 
As a consequence, when someone chooses `Option 2`, the probability of this choice given our model would be $0$. 
In contrast, if instead $\theta=.3$, the probability of someone choosing `Option 2` would be instead $.7$. 
In other words, the observed choice of `Option B` is more likely when $\theta = .3$ than when $\theta=1$. 
Such a model of course disregards any attributes (outcome values and probabilities) of a given choice problem.

```{r}
# theta=1
theta <- 1
B_choice <- 0 
dbinom(B_choice,1,prob=theta)

# theta = .3
theta <- .3
B_choice <- 0
dbinom(B_choice,1,prob=theta)
```

We can now extend this example to more choices and more candidate values for $\theta$. 
Note that the likelihood for multiple choices is simply the product of all choices' individual likelihoods, assuming that the choices are independent.
In the following example, we take the candidate values `\theta=``r seq(0,1,.1)` and calculate the likelihood of the choices of all participants on Problem 1 from Kellen et al. (2016) data given these parameter values. 

```{r}
# get choices from problem 1
choices_p1 <- as.numeric(as.vector(cpt_choices[1,]))
choices_p1

# calculate the likelihood
theta <- seq(0,1,.1)
lh <- round(dbinom(sum(choices_p1),length(choices_p1), prob=theta), 3)
tibble(Theta=theta , 
       Likelihood = lh, 
       Likelihood_std = lh/sum(lh)) %>%
  kable()
```

We can easily see that the observed choices on Problem 1 are much more likely given the data when we assume a moderate to high probability of choosing `Option 1` (parameter values between 0.7 and 0.8) when we rely on this simple binomial decision model.

#### Likelihood for CPT

Although CPT is a more complex model than the simple binomial from above, the idea is exactly the same. 
We have a number of free (unknown) parameters that we want to learn about ($\theta$ for the binomial; $\alpha, \lambda,\gamma,\delta$ for CPT) and the probability of observed choices depends on the values of these parameters.
The main difference between the binomial and CPT is that the binomial simply assumes a constant probability for choosing `Option 1` vs. `Option 2` for all choice problems (that is $\theta$), whereas CPT uses (transformations of) the outcomes and their probabilities in a choice problem to compute a choice probability that is specific for this problem. 
However, since CPT can only make deterministic predictions so far (e.g., always choose `Option A` when $V_A > V_B$), we need to make one more adjustment to CPT so that it can provide us with probabilities that can also be different from 0 and 1.

In this adjustment, we use a choice rule, which predicts a higher probability for choosing `Option A`, the better its CPT score ($V_A$) is compared to the score for `Option B`. 
That is, in contrast to the deterministic version, the prediction does not only depend on whether `Option A` obtains a better evaluation than `Option B`, but also on how much better this evaluation is. 
If the evaluation is much better, it is highly likely that `Option A` is chosen, if it is much worse, it is instead highly likely that `Option B` is chosen. 
If their evaluations are similar, the choice probability gets closer to $.5$.
This logic is implemented in the so called logit choice rule, which takes the CPT valuations as input and returns a choice probability as output: 
$$
p(A) = \frac{1}{1+e^{-\phi(V_A-V_B)}} \: , 
$$

where $\phi$ is a number that can take values $\geq0$ (free parameter).
The figure below illustrates how the logit choice rule translates differences in the CPT valuations (x-axis) into a choice probability that is constrained between 0 and 1 (y-axis).
In general, when difference in valuations is 0, the choice probability is .5. 
Intuitively, when the options do not really differ, we also have no preference for one option or the other. 
Then, when the difference in valuations becomes positive, the probability for choosing `Option A` increases; and when they become negative, the probability for choosing `Option A` decreases.
As can be seen, $\phi$ affects how strongly the probability depends on the differences. 

```{r}
logit <- expand_grid(Diff = c(seq(-4, 4, .01)) , 
                     phi = c(0, .5, 1)) %>%
  mutate(Prob = 1/(1+(exp(-phi*Diff)) ))


logit %>% ggplot(aes(Diff, Prob, group = as.factor(phi), color = as.factor(phi))) +
  geom_line(linewidth = 1) + 
  theme_minimal()
```

## Using MCMC/JAGS to obtain the posterior

```{r eval=FALSE}

nGain <- cpt_problems %>% filter(domain=='Gain') %>% nrow()
nLoss <- cpt_problems %>%  filter(domain=='Loss') %>% nrow()
nTotal <- cpt_problems %>% nrow()
nSubj <- ncol(cpt_choices)


params_cpt <- c("alpha", "gamma", "delta", "rho") # free parameters
#source("code/helper_functions/fun_initialize_MCMC.R") # calls function to create starting values for MCMC

params_init <- function(){
  list("alpha.pre" = rep(.5, nSubj) , # individual level parameters
       "gamma.pre" = rep(.5, nSubj) ,
       "delta.pre" = rep(.5, nSubj) , 
       "rho.pre" = rep(.001, nSubj) 
       ) 
}


data <- list(choice = as.matrix(cpt_choices) ,
             hA = cpt_problems$hA ,
             lA = cpt_problems$lA ,
             hB = cpt_problems$hB ,
             lB = cpt_problems$lB ,
             p_hA = cpt_problems$p_hA ,
             p_lA = cpt_problems$p_lA ,
             p_hB = cpt_problems$p_hB ,
             p_lB = cpt_problems$p_lB ,
             nGain = nGain , 
             nLoss = nLoss , 
             nTotal = nTotal , 
             nSubj = nSubj
  )

  ## sample from posterior distributions using MCMC
m <- jags.parallel(data = data , 
                   inits = params_init , 
                   parameters.to.save = params_cpt ,
                   model.file = "tutorials/BayesianCognitiveModeling/JAGS_cpt_model_independent.txt" # see below ,
                   n.chains = 6,
                   n.iter = 2000 ,
                   n.burnin = 1000 ,
                   n.cluster = 6 , # compute MCMC chains in parallel
                   DIC = TRUE ,
                   jags.seed = 1223)
```

```{r}
#| eval: false
#| echo: false

saveRDS(m, 'tutorials/BayesianCognitiveModeling/cpt_model.rds')
```


```
# JAGS_cpt_model_independent.txt


model {

  for (j in 1:nSubj) {  #Subject-loop

    #Value function 
    alpha.pre[j] ~ dbeta(1,1)
    gamma.pre[j] ~ dbeta(1,1)
    delta.pre[j] ~ dbeta(1,1)
    rho.pre[j] ~ dbeta(1,1) 
    
    alpha[j] <- (alpha.pre[j])
    gamma[j] <- (gamma.pre[j]) * 2
    delta[j] <- (delta.pre[j]) * 5
    rho[j] <- (rho.pre[j]) * 5 
  

    #Gains
    for (i in 1:nGain){  #Item loop

    
      #Value function
      #Lottery A
      v.hA[i,j] <- pow(hA[i],alpha[j])
      v.lA[i,j] <- pow(lA[i],alpha[j])
      
      #Lottery B
      v.hB[i,j] <- pow(hA[i],alpha[j])       
      v.lB[i,j] <- pow(hB[i],alpha[j])
      
      #Probability weighting function 
      #Lottery A
      pi.hA[i,j] <- (delta[j] * (pow(p_hA[i], gamma[j]))) / (delta[j] * (pow(p_hA[i], gamma[j])) + pow(p_lA[i], gamma[j]))
      pi.lA[i,j]<- 1 - pi.hA[i,j]
      
      #Lottery B
      pi.hB[i,j] <- (delta[j] * (pow(p_hB[i], gamma[j]))) / (delta[j] * (pow(p_hB[i], gamma[j])) + pow(p_lB[i], gamma[j]))
      pi.lB[i,j] <- 1 - pi.hB[i,j]
      
      #Valuation 
      Vf.A[i,j]  <- pi.hA[i,j] * v.hA[i,j] + pi.lA[i,j] * v.lA[i,j]
      Vf.B[i,j]  <- pi.hB[i,j] * v.hB[i,j] + pi.lB[i,j] * v.lB[i,j]
      
      #Choice rule
      binval[i,j] <- (1)/(1+exp(-1*(rho[j]*(Vf.A[i,j]-Vf.B[i,j]))))
      choice[i,j] ~ dbern(binval[i,j])
      
    }
    
    
    
    # Losses
    for (i in (nGain+1):(nGain+nLoss)) {
    
    
      #Value function
      #Lottery A
      v.hA[i,j] <- (-1) * pow((-1*hA[i]),alpha[j])
      v.lA[i,j] <- (-1) * pow((-1*lA[i]),alpha[j])
  
      #Lottery B
      v.hB[i,j] <- (-1) * pow((-1*hB[i]),alpha[j])     
      v.lB[i,j] <- (-1) * pow((-1*hB[i]),alpha[j])
      
      #Probability weighting function 
      #Lottery A
      pi.lA[i,j] <- (delta[j] * (pow(p_lA[i], gamma[j]))) / (delta[j] * (pow(p_lA[i], gamma[j])) + pow(p_hA[i], gamma[j]))
      pi.hA[i,j] <- 1 - pi.lA[i,j]
  
      #Lottery B
      pi.lB[i,j] <- (delta[j] * (pow(p_lB[i], gamma[j]))) / (delta[j] * (pow(p_lB[i], gamma[j])) + pow(p_hB[i], gamma[j]))
      pi.hB[i,j] <- 1 - pi.lB[i,j]
  
      #Valuation 
      Vf.A[i,j]  <- pi.hA[i,j] * v.hA[i,j] + pi.lA[i,j] * v.lA[i,j]
      Vf.B[i,j]  <- pi.hB[i,j] * v.hB[i,j] + pi.lB[i,j] * v.lB[i,j]
  
      #Choice rule
      binval[i,j] <- (1)/(1+exp(-1*(rho[j]*(Vf.A[i,j]-Vf.B[i,j]))))
      choice[i,j] ~ dbern(binval[i,j])
      
      }
      
      
      # Mixed
       
      for (i in (nGain+nLoss+1):nTotal) {
    
    
      #Value function
      #Lottery A
      v.hA[i,j] <- (-1) * pow(hA[i],alpha[j])
      v.lA[i,j] <- (-1) * pow((-1*lA[i]),alpha[j])
  
      #Lottery B
      v.hB[i,j] <- (-1) * pow(hB[i],alpha[j])   
      v.lB[i,j] <- (-1) * pow((-1*lB[i]),alpha[j])
  
      #Probability weighting function 
      #Lottery A
      pi.hA[i,j] <- (delta[j] * (pow(p_hA[i], gamma[j]))) / (delta[j] * (pow(p_hA[i], gamma[j])) + pow(p_lA[i], gamma[j]))
      pi.lA[i,j]<- 1 - pi.hA[i,j]
      
      #Lottery B
      pi.hB[i,j] <- (delta[j] * (pow(p_hB[i], gamma[j]))) / (delta[j] * (pow(p_hB[i], gamma[j])) + pow(p_lB[i], gamma[j]))
      pi.lB[i,j] <- 1 - pi.hB[i,j]
  
      #Valuation 
      Vf.A[i,j]  <- pi.hA[i,j] * v.hA[i,j] + pi.lA[i,j] * v.lA[i,j]
      Vf.B[i,j]  <- pi.hB[i,j] * v.hB[i,j] + pi.lB[i,j] * v.lB[i,j]
  
      #Choice rule
      binval[i,j] <- (1)/(1+exp(-1*(rho[j]*(Vf.A[i,j]-Vf.B[i,j]))))
      choice[i,j] ~ dbern(binval[i,j])
      
      }
      
      
      
      
  }
}   

```

```{r}
#| echo: false

m <- readRDS(here('tutorials', 'BayesianCognitiveModeling', 'cpt_model.rds'))
```


```{r}
m.summary <- m$BUGSoutput$summary %>% as_tibble(rownames = "parameter")
kable(m.summary, digits = 4)
```



